---
title: "Working with Tables"
date: "2022-10-26"
output:
  html_document:
    highlight: tango
    number_sections: yes
    theme: readable
    toc: yes
    toc_float: yes
---

# Goals for today:

1. Practice Making Publication Quality Tables with `flextable` library
2. Create a theme for your `ggplot` visualizations
3. Creating Animations/Interactive Plots
4. If time permits, explore some html features

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(DT)
library(flextable)
library(ggplot2)
library(gganimate)
library(gapminder)
library(plotly)
library(cowplot)
library(transformr)
```

# Making Nice Tables


## Task 1:  

Take `diamonds` dataset, select columns `cut`, `color`, `price`, and `depth`. Filter the data so that dataset only have diamonds with cut quality Fair and Good. Then, group dataset by `cut` and `color` and produce a summary the mean and standard deviation for depth and price. Assign name `data_color` to your final output.

```{r}
data_color <- diamonds %>% 
  select(cut, color, price, depth) %>% 
  filter(cut %in% c("Fair", "Good")) %>% 
  group_by(cut, color) %>% 
  summarise(mean_depth = mean(depth), sd_depth = sd(depth), mean_price = mean(price), sd_price = sd(price), .groups = "keep")

#.groups = "keep" ensures the groups specified above, remain
# can althernatively use the NEW pipe |>

print(data_color)
```


## Viewing Data


More info about `DT` library can be found here: https://rstudio.github.io/DT/
```{r}
datatable(data_color)

#datatable package, DT, creates an interactive html type of table. You can filter by search!

```

## Producing Publication Quality Tables

The user manual is available here: https://ardata-fr.github.io/flextable-book/ with list of all available functions: https://davidgohel.github.io/flextable/reference/index.html

```{r}
#flextable() produces formattable tables. Default format is the ___? (displayed below)
tbl<-  data_color |>  flextable() |>
  #we can use colformat and digits specification to limit decimal points
  colformat_double(digits = 1) %>% 
  #linespacing function enables us to change the spacing between obesrvations in the table
  line_spacing(space = 0.5) %>% 
  #set_caption allows us to add a title
  set_caption("Summary Statistics") |>
  #the flextable merge function is useful for collapsing repetitive things into 1. note  j = columns, i = rows
  merge_v(j=c("cut")) %>% 
  #valign lets us move the merged labels around. part = body refers to only the body of the table is where we want to align. valign = top means we want stuff to be at the top
  valign(j=1, valign = "top", part = "body") %>% 
  #below, using align_nottext_col to align to center of titles in each column.
  align_nottext_col(align = "center") %>% 
  # j means columns. We're bolding the first and second columns in the body only
  bold(j=c(1:2), bold = TRUE, part = "body") %>% 
  #with only the above calls, we have a problem with the line on the bottom. We can use a theme to fix this
  theme_booktabs() %>% 
  #autofit (below), ensures anything we produce after will always fit in the boundaries of the page
  autofit()

tbl

#the below command opens the tibble in microsoft word as a figure
#print(tbl, "docx")
# save as docx will let you save the table
#save_as_docx()
```

```{r}
typology <- data.frame(
  col_keys = colnames(data_color),
  #c("cut", "color", "price_mean", "price_sd", "depth_mean", "depth_sd"
  names = c("Cut","Color","Depth","Depth","Price","Price"),    # this is the top header
  metric = c("Cut","Color", "mean", 'sd', "mean", 'sd'),    # this is the second header
  stringsAsFactors = FALSE)

tbl |> 
  set_header_df(mapping = typology, key = "col_keys" ) %>% 
  theme_booktabs() %>% 
  align(align = "center", part = "all") %>% 
  merge_h(part = "header") %>% 
  merge_v(part = "header") %>% 
  fix_border_issues(part = "all") %>% 
  vline(j=c(4))

```



STOPPED HERE IN CLASS



# Publication Quality Theme

The function created by Koundinya Desiraju https://rpubs.com/Koundy/71792

```{r}
# library(extrafont)
# loadfonts(quiet = TRUE)
# fonts()
theme_Publication <- function(base_size=14, base_family="Times New Roman") {
      library(grid)
      library(ggthemes)
    (theme_foundation(base_size=base_size, base_family=base_family)
    + theme(plot.title = element_text(face = "bold",
                                      size = rel(1.2), hjust = 0.5),
            text = element_text(),
               panel.background = element_rect(colour = NA),
               plot.background = element_rect(colour = NA),
               panel.border = element_rect(colour = NA),
               axis.title = element_text(face = "bold",size = rel(1)),
               axis.title.y = element_text(angle=90,vjust =2),
               axis.title.x = element_text(vjust = -0.2),
               axis.text = element_text(),
               axis.line = element_line(colour="black"),
               axis.ticks = element_line(),
               panel.grid.major = element_line(colour="#f0f0f0"),
               panel.grid.minor = element_blank(),
               legend.key = element_rect(colour = NA),
               legend.position = "right",
               legend.direction = "vertical",
               legend.key.size= unit(0.2, "cm"),
               legend.spacing = unit(0, "cm"),
               legend.title = element_text(face="italic"),
               plot.margin=unit(c(10,5,5,5),"mm"),
               strip.background=element_rect(colour="#f0f0f0",fill="#f0f0f0"),
               strip.text = element_text(face="bold")
          ))

}

scale_fill_Publication <- function(...){
      library(scales)
      discrete_scale("fill","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}

scale_colour_Publication <- function(...){
      library(scales)
      discrete_scale("colour","Publication",manual_pal(values = c("#386cb0","#fdb462","#7fc97f","#ef3b2c","#662506","#a6cee3","#fb9a99","#984ea3","#ffff33")), ...)

}
```

```{r warning=FALSE}
ggplot(mtcars, aes(mpg, disp, color = factor(carb))) +
  geom_point(size = 3) + labs(title = "Scatter Plot", color="carb")
```

```{r warning=FALSE}
ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar(alpha=0.9) + 
  labs(title="Bar Plot", color="carb")

```

```{r warning=FALSE}
ggplot(data = iris,aes(Sepal.Length, Petal.Length,color=Species))+
  geom_point(size=3) + 
  facet_wrap(~ Species,scales="free")+
  labs(title="Plot With Facets")
```


## Task 2:

Using `gapminder` dataset,  produce a geom_point() plot for `gdpPercap` and `lifeExp`. Colour the points by `country`, make size of the points be proportional to the population.  Make it look publishable.

```{r}
gap.plt<-

gap.plt
```

## Making Plots Interactive

```{r}
ggplotly(gap.plt, tooltip = c("country","x","y","pop"))
```

## Animations

Animations or GIFs (Graphics Interchange Format) can be generated using the `gganimate` package written by Thomas Lin Pedersen and David Robinson. The `gganimate` package allows to track changes over time while simultaneously displaying several variables in one visualization. 

```{r}
gif <- gap.plt + transition_time(year) +
  labs(title = "Year: {frame_time}")
# show gif
gif
```

```{r warning=FALSE}
fig <- gapminder %>%
  plot_ly(
    x = ~gdpPercap, 
    y = ~lifeExp, 
    size = ~pop, 
    color = ~continent, 
    frame = ~year, 
    text = ~country, 
    hoverinfo = "text",
    type = 'scatter',
    mode = 'markers'
  )
fig <- fig %>% layout(
    xaxis = list(
      type = "log"
    )
  )

fig
```








