---
title: "Bullet_Data_Analysis"
author: "Matt"
date: "2022-10-31"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(rstatix)
library(tidyverse)
library(scales)
library(summarytools)
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```


Loading the data: 

```{r}
df <- read_csv("https://github.com/firasm/bits/raw/master/bullet_data.csv")

head(df)
```

Getting summary statistics using two different methods:

```{r}
#method 1
psych::describe(df)

#method 2
df %>% get_summary_stats(type="full", show=c("n", "mean","median","sd","min", "q1", "q3", "max"))
```

Another method to get data summaries:

```{r}
stview(dfSummary(df),  plain.ascii=FALSE)
```



converting zones to factor type:

```{r}
df$zone=as.factor(df$zone)
```



removing non-hit observations:

```{r}
hits_df<- df %>% filter(bullet==1) 

head(hits_df)
```


counting the number of hits in each zone

```{r}
summary <- hits_df %>% group_by(zone) %>% count() 

summary
```

```{r}
hits_plot <- hits_df %>% 
  ggplot() +   
  geom_histogram(aes(y=zone), stat="count", fill="steelblue") +   
  labs(x="Bullet hits", y="", title = "Bullet hit count by Airplane Zone") +  
  scale_y_discrete(limits=rev,drop=FALSE,expand = c(0,0)) + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  theme_classic()

hits_plot
```

This is very helpful, but not quite done yet. Tihis is still summary data, we can guess that the two areas may need to be reinforced are B and E, but lets go a bit further...

```{r}
outline <- df %>% filter(zone=="OutsidePlane") %>% ggplot()+   geom_point(aes(y,-x, color="OutsidePlane"))+   labs(color="bullet")+   theme_void()

outline
```

```{r}
outline + 
  geom_point(data=hits_df, aes(y,-x,color=zone),alpha=0.8, inherit.aes = FALSE)
```


Interestingly, there are areas that have no bullet holes ---- these areas may atually be the fatal areas (remember we're working with the planes that DID COME BACK) - we dont know that these areas are fatal for sure, but we can now investigate this and understand. 







