---
title: "MEDI 504A - Lab 3"
author: "Matthew Manson"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

# MEDI 504 Lab 3

## Basic biostatistics

By the end of this lab, students should be able to:

-   Identify the different types of data analysis questions and
    categorize a question into the correct type
-   Identify a suitable analysis type to answer an inferential question,
    given the data set at hand
-   Use the R programming language to carry out analysis to answer
    inferential question
-   Interpret and communicate the results of the analysis from an
    inferential question

```{r include=FALSE}
library(tidyverse)
library(broom)
library(binom)
```

## Exercise 1: types of data analysis questions

rubric={10 points}

In the reading [**Types of data analytic
questions**](https://jtleek.com/ads2020/week-1.html#types-of-data-analytic-questions),
you were introduced to different types of statistical questions. Let us
refresh our knowledge of these here and play name that statistical
question! For each question below, assign the answer to one of the
following types of statistical question being asked:

-   **Descriptive.**
-   **Exploratory.**
-   **Inferential.**
-   **Predictive.**
-   **Causal.**
-   **Mechanistic.**

### Exercise 1.1:

Is wearing sunscreen associated with a decreased probability of
developing skin cancer in Canada?

> Inferential

### Exercise 1.2:

Is there a relationship between alcohol consumption and socioeconomic
status in the 2018 City of Vancouver survey data set?

> Exploratory

### Exercise 1.3:

Does performing strength training 3 times a week lead to an increase in
bone density in the elderly?

> Causal

### Exercise 1.4:

How do changes in human behaviour lead to a reduction in the number of
COVID-19 confirmed cases?

> Mechanistic???

### Exercise 1.5:

Does reduced caloric intake cause weight-loss in adults?

> Causal

### Exercise 1.6:

Do countries with lower COVID-19 vaccination rates have higher levels of
hospitalizations compared to countries with higher COVID-19 vaccination
rates?

> Inferential

### Exercise 1.7:

Is vaccination against COVID-19 negatively associated with the presence
long-COVID symptoms?

> Inferential

### Exercise 1.8:

How many patients will go to the emergency department at Vancouver
General Hospital tomorrow?

> Predictive

### Exercise 1.9:

How many COVID-19 patients are in BC hospitals today?

> Descriptive

### Exercise 1.10:

Are high contrast images associated with better visual discrimination by
the visually impaired?

> Inferential




## Exercise 2: identifying a suitable analysis method for a given question and data set

rubric={10 points}

Given the statistical question below and the data set description and
snippet, name the type of statistical question and a suitable analysis
method. Justify your choices for both the question type and analysis
method. *Note: this case is fictional, but based on available medical
methods.*

**Statistical question:** Is there a difference in the proportion of
miscarriages in in-vitro fertilization (IVF) patients whose embryos
undergoe preimplantation genetic testing for aneuploidy (PGT) compared
to those whose embryos do not?

> Question Type: Inferential
> Justification: This question suggests that the results of the analysis will be generalized to all individuals who undergo IVF. For this reason this question can be considered an inferential statistical question - it goes beyond the population from which the data was generated.                                            

**Data set:** Data from 457 patients was collected from the a local
fertility clinic. Patients had the choice of opting for PGT or not.
There is an added financial cost for PGT and therefore not all patents
chose to opt for this added treatment. 196 patients opted to undergo PGT
screening of their embryos, and 261 opted to forgo this screening.
Miscarriage proportion for each patient was calculated as the number of
unsuccessful embryo transfers divided by the total number of embryo
transfers (successful + unsuccessful). A snippet of the data is shown
below:

| `patient_id` | `miscarriage_proportion` | `pgt` |
|:-------------|:-------------------------|:------|
| 2361344      | 0.25                     | yes   |
| 2361932      | 0.33                     | no    |
| 2397563      | 0                        | no    |
| ...          | ...                      | ...   |
| 2595244      | 1                        | yes   |

> Analysis Method: A method to compare two proportions - z-test, or Chi-square test. 
> Justification: Here, the researchers are trying to compare the *proportion* of miscarriages experienced by two groups. When comparing *proportions*, we can use one of the two methods above, however their assumptions differ. 




## Exercise 3: using R to visualize uncertainty of point estimates

rubric={20 points}

In a recent study by Jiang et al (2019), they investigated the effects
of intramuscular and vaginal progesterone supplementation on
frozen-thawed embryo transfer during in-vitro fertilization (IVF). This
is an important question because progesterone supplementation is
critical during IVF frozen-thawed embryo transfer, and intramuscular
supplementation has many negative side effects (e.g., inconvenience,
local pain and inflammation at the injection site). Patients were
assigned to one of two groups: - group A with progesterone intramuscular
injection (60 mg/d) - group B with progesterone vaginal
sustained-release gel of progesterone (90 mg/d)

The response variable of interest was whether a pregnancy resulted in a
live birth (coded as 1) or not (coded as 0).

Your task here is to load the `data/jiang-live-birth.csv` file and
create an effective data visualization which communicates the estimates
for each group (proportion of live births) as well as the uncertainty of
those point estimates at a 95% confidence interval.

```{r, fig.width = 6, fig.height = 5}

#load the data
jiang_live_birth <- read.csv("data/jiang-live-birth.csv") 

#a look at the data, as imported
jiang_live_birth 

#creating new data frame displaying the number of live birth 
#successes, total attempts, and proportion of successes for each group
(jiang_live_birth1 <- jiang_live_birth %>% 
  group_by(group) %>% 
  summarise(live_births = sum(live_birth), total = n()) %>% 
  mutate(prop = live_births/total))

#95% confidence interval for each group
confidence <- binom.confint(jiang_live_birth1$live_births, jiang_live_birth1$total, conf.level = 0.95, methods = "exact") 

#binding column to jiang_live_birth_1
jiang_live_birth_final <- cbind(jiang_live_birth1, confidence$lower, confidence$upper)

(jiang_live_birth_final <- jiang_live_birth_final %>% rename(lower = 'confidence$lower', upper = 'confidence$upper'))

#plotting proportions of live births in each group
jiang_live_birth_final %>% 
ggplot(aes(x = group, y = prop)) +
  geom_col() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "The Proportion of Live Birth Successes in each Progesterone Supplementation Method Group",
       x = "Progesterone Supplementation Method", 
       y = "Proportion of Live Birth Successes")



### NOT DONE ###
  


```




## Exercise 4: using R to infer group differences

rubric={20 points}

The error bars representing the uncertainty of our estimates in the
visualization overlap! From this visualization alone, it is not yet
clear as to whether the observed difference in the estimates is
statistically significant. Perform a suitable analysis to answer this
question. If any hypotheses or assumptions are made in your analysis,
state them. Clearly communicate your results.

```{r}
# Your code goes here

#p value alone is no enough - effect size, confidence interval, etc. 


```

> Your answer goes here.

## (Optional) Exercise 5: using R to handle multiple comparisons

rubric={3 bonus points}

We will be working with the results from a Genome-wide analysis-like
study found in `data/GWAS_results` from Timbers et al. (2016). The
dataset contains two columns: a list of gene names (`gene`) and a list
of unadjusted $p$-values (`pval`) generated from the analysis (the
particular statistical test used is the Sequence Kernel Analysis test).
These $p$-values were created by repeating the analysis on many
variables from the same dataset. Thus we have a multiple testing problem
to deal with. Each $p$-value corresponds to a gene and tests whether
that gene is associated with a phenotype.

Note: before you get started on this question we recommend you read the
following: - [Types of errors section of the Modern Dive statistics
textbook](https://moderndive.com/9-hypothesis-testing.html#types-of-errors) -
[Why is multiple testing a problem and what do I need to do about it?
slides](https://github.com/ttimbers/SFU_omics_2018/blob/master/timbers_SFU_omics_2018.pdf)

```{r}
GWAS_results <- read_csv("data/GWAS_results.csv", show_col_types = FALSE) %>%
  select(gene = public_gene_name, pval = `p-value`)
GWAS_results
```

Using in a sample of 480 mutant C. elegans (nematode worms), the
question in the analysis was: **are there any genes, which when mutated,
associated with a phenotype defined as a decrease in the ability to
uptake a fluorescent dye into their sensory neurons?** This would
indicate there might be a problem with their sensory neurons and that
this gene might be important for sensory neuron development or function.
The study can be accessed here:
<http://dx.doi.org/10.1371/journal.pgen.1006235>

## (Optional) Exercise 5.1:

Answer the following questions

-   How many genes are present in the total dataset? For this dataset,
    this corresponds to the number of multiple comparisons that were
    performed.
-   How many genes are associated with the phenotype (a decrease in the
    ability to uptake a fluorescent dye into their sensory neurons) at
    the unadjusted $\alpha = 0.05$?

```{r}
# Your code goes here

```

> Your answer goes here.

## (Optional) Exercise 5.2:

Briefly describe (in one or two sentences) why it would be misleading to
report only one of the "significant" tests (and ignoring the fact that
others were done too).

> Your answer goes here.

## (Optional) Exercise 5.3:

Use the function `p.adjust()` to calculate adjusted $p$-values using
`method = "bonferroni"`. How many and which genes are associated with
the treatment after the adjustment, at the $\alpha = 0.05$ significance
level?

```{r}
# Your code goes here

```

> Your answer goes here.

## References

Jiang, L., Luo, ZY., Hao, GM. et al. Effects of intramuscular and
vaginal progesterone supplementation on frozen-thawed embryo transfer.
Sci Rep 9, 15264 (2019). <https://doi.org/10.1038/s41598-019-51717-5>

Timbers TA, Garland SJ, Mohan S, Flibotte S, Edgley M, et al. (2016)
Accelerating Gene Discovery by Phenotyping Whole-Genome Sequenced
Multi-mutation Strains and Using the Sequence Kernel Association Test
(SKAT). PLOS Genetics 12(8): e1006235.
<https://doi.org/10.1371/journal.pgen.1006235>
